---
bibliography: references.bib
---

```{r}
dist_loc <- list.files(
  find.package("DiagrammeR"),
  recursive = TRUE,
  pattern = "mermaid.*js",
  full.names = TRUE
)
js_cdn_url <- "https://cdnjs.cloudflare.com/ajax/libs/mermaid/9.0.1/mermaid.min.js"
download.file(js_cdn_url, dist_loc)
```

# Workflow

You could check this book for metabolomics data analysis [@li2020a].

```{r}
DiagrammeR::mermaid("
flowchart TB
I(peak-picking) --> C
C(visulization) --> D(normalization/batch correction)
D --> A(annotation/identification)
A --> H(statistical analysis)
C --> A --> B(omics analysis)
D --> H
B --> H
H --> E(experimental validation)
A --> E
H --> A
B --> E
C --> H
")
```

## Platform for metabolomics data analysis

Here is a list for related open source [projects](http://strimmerlab.org/notes/mass-spectrometry.html)

### XCMS & XCMS online

[XCMS online](https://xcmsonline.scripps.edu/landing_page.php?pgcontent=mainPage) is hosted by Scripps Institute. If your datasets are not large, XCMS online would be the best option for you. Recently they updated the online version to support more functions for systems biology. They use metlin and iso metlin to annotate the MS/MS data. Pathway analysis is also supported. Besides, to accelerate the process, xcms online employed stream (windows only). You could use stream to connect your instrument workstation to their server and process the data along with the data acquisition automate. They also developed apps for xcms online, but I think apps for slack would be even cooler to control the data processing.

[xcms](https://bioconductor.org/packages/release/bioc/html/xcms.html) is different from xcms online while they might share the same code. I used it almost every data to run local metabolomics data analysis. Recently, they will change their version to xcms 3 with major update for object class. Their data format would integrate into the MSnbase package and the parameters would be easy to set up for each step. Normally, I will use msconvert-IPO-xcms-xMSannotator-metaboanalyst as workflow to process the offline data. It could accelerate the process by parallel processing. However, if you are not familiar with R, you would better to choose some software below. For xcms, 1000 files will need around 5 hours to generate the peaks list on a regular workstation.

[IPO](https://github.com/rietho/IPO) A Tool for automated Optimization of XCMS Parameters [@libiseller2015] and [Warpgroup](https://github.com/nathaniel-mahieu/warpgroup) is used for chromatogram subregion detection, consensus integration bound determination and accurate missing value integration[@mahieu2016]. A case study to compare different xcms parameters with IPO can be found for GC-MS [@dossantos2023]. Another option is AutoTuner, which are much faster than IPO[@mclean2020]. Recently, MetaboAnalystR 3.0 could also optimize the parameters for xcms while you need to perform the following analysis within this software[@pang2020]. For IPO, ten files will need \~12 hours to generate the optimized results on a regular workstation. [Paramounter](https://github.com/HuanLab/Paramounter) is a direct measurement of universal parameters to process metabolomics data in a “White Box”[@guo2022]. Another research use machine learning method to compare different optimization methods and they are all better than the default setting of xcms[@lassen2021]. It could be extended to include ion mobility[@dodds2022].

Check those papers for the XCMS based workflow[@forsberg2018; @huan2017; @mahieu2016a; @montenegro-burke2017; @domingo-almenara2020; @stancliffe2022]. For metlin related annotation, check those papers[@guijas2018; @tautenhahn2012; @xue2020; @domingo-almenara2018a].

[MAIT](https://www.bioconductor.org/packages/release/bioc/html/MAIT.html) based on xcms and you could find source code [here](https://github.com/jpgroup/MAIT)[@fernandez-albert2014a].

[iMet-Q](http://ms.iis.sinica.edu.tw/comics/Software_iMet-Q.html) is an automated tool with friendly user interfaces for quantifying metabolites in full-scan liquid chromatography-mass spectrometry (LC-MS) data [@chang2016]

compMS2Miner is an Automatable Metabolite Identification, Visualization, and Data-Sharing R Package for High-Resolution LC--MS Data Sets. Here is related papers [@edmands2017; @edmands2018; @edmands2015].

mzMatch is a modular, open source and platform independent data processing pipeline for metabolomics LC/MS data written in the Java language, which could be coupled with xcms [@scheltema2011; @creek2012]. It also could be used for annotation with MetAssign[@daly2014].

### PRIMe

[PRIMe](http://prime.psc.riken.jp/Metabolomics_Software/) is from RIKEN and UC Davis. They update their database frequently[@tsugawa2016]. It supports mzML and major MS vendor formats. They defined own file format ABF and eco-system for omics studies. The software are updated almost everyday. You could use MS-DIAL for untargeted analysis and MRMPROBS for targeted analysis. For annotation, they developed MS-FINDER and statistic tools with excel. This platform could replace the expensive software from vendors and is well prepared for MS/MS data analysis and lipidomics. They are open source, work on Windows and also could run within mathematics. However, they don't cover pathway analysis. Another feature is they always show the most recently spectral records from public repositories. You could always get the updated MSP spectra files for your own data analysis.

MS-DIAL 4 added support for lipidomics with an integrated CCS and retention time atlas[@tsugawa2020]. The latest version, MS-DIAL 5, further extends the platform with multimodal mass spectrometry data mining capabilities including improved DIA deconvolution[@tsugawa2024].

For PRIMe based workflow, check those papers[@lai2018; @matsuo2017; @treutler2016; @tsugawa2015; @tsugawa2016; @kind2018]. There are also extensions for their workflow[@uchino2022] and workflow for environmental science[@bonnefille2023].

### GNPS

[GNPS](https://gnps.ucsd.edu) is an open-access knowledge base for community-wide organization and sharing of raw, processed or identified tandem mass (MS/MS) spectrometry data. It's a straightforward annotation method for MS/MS data. Feature-based molecular networking (FBMN) within GNPS could be coupled with xcms, OpenMS, MS-DIAL, MZmine, and other popular software. GNPS also has a dashboard for online mass spectrometry data analysis[@petras2021a]. Ion identity molecular networking[@schmid2021a] further extends GNPS by connecting features from the same compound (adducts, in-source fragments) into ion identity networks, improving both annotation accuracy and molecular networking results.

Check those papers for GNPS and related projects[@aron2020; @nothias2020; @scheubert2017; @silva2018; @wang2016b; @bittremieux2023; @schmid2021a].

### OpenMS & SIRIUS

[OpenMS](https://www.openms.de/) is another good platform for mass spectrum data analysis developed with C++. You could use them as plugin of [KNIME](https://www.knime.org/). I suggest anyone who want to be a data scientist to get familiar with platform like KNIME because they supplied various API for different programme language, which is easy to use and show every steps for others. Also TOPPView in OpenMS could be the best software to visualize the MS data. You could always use the metabolomics workflow to train starter about details in data processing. pyOpenMS and OpenSWATH are also used in this platform. If you want to turn into industry, this platform fit you best because you might get a clear idea about solution and workflow.

Check those paper for OpenMS based workflow[@bertsch2011; @pfeuffer2017; @rost2014; @rost2016; @rurik2020; @alka2020;@pfeuffer2024].

OpenMS could be coupled to SIRIUS 4 for annotation. [Sirius](https://bio.informatik.uni-jena.de/software/sirius/) is a new java-based software framework for discovering a landscape of de-novo identification of metabolites using single and tandem mass spectrometry. SIRIUS 4 project integrates a collection of our tools, including [CSI:FingerID](https://www.csi-fingerid.uni-jena.de/), [ZODIAC](https://bio.informatik.uni-jena.de/software/zodiac/) and [CANOPUS](https://bio.informatik.uni-jena.de/software/canopus/). Check those papers for SIRIUS based workflow[@duhrkop2019; @duhrkop2020; @alka2020; @ludwig2020].

### MZmine

[MZmine](https://mzmine.github.io/) was originally developed on the Java platform. MZmine 2[@pluskal2010] was widely used in the community. In 2023, MZmine 3 was released with a completely rewritten architecture, adding support for ion mobility spectrometry data, improved feature detection, and native integration with GNPS molecular networking and SIRIUS[@schmid2023]. MZmine 3 is now the recommended version and is one of the most actively maintained open-source platforms for untargeted metabolomics. However, MZmine does not have built-in pathway analysis. You could use MetaboAnalyst for that purpose.

Check those papers for MZmine based workflow[@pluskal2010; @pluskal2020; @schmid2023].

### Emory MaHPIC

This platform is composed by several R packages from Emory University including [apLCMS](https://sourceforge.net/projects/aplcms/) to collect the data, [xMSanalyzer](https://sourceforge.net/projects/xmsanalyzer/) to handle automated pipeline for large-scale, non-targeted metabolomics data, [xMSannotator](https://sourceforge.net/projects/xmsannotator/) for annotation of LC-MS data and Mummichog for pathway and network analysis for high-throughput metabolomics. Note that the original Mummichog is no longer actively maintained; its algorithm is now integrated into [MetaboAnalyst](https://www.metaboanalyst.ca/)[@pang2024]. This platform would be preferred by someone from environmental science to study exposome.

You could check those papers for Emory workflow[@uppal2013; @uppal2017; @yu2009b; @li2013; @liu2020].

### Others

- [MetaboAnalyst](https://www.metaboanalyst.ca/) is a comprehensive web-based platform for metabolomics data analysis, covering statistical analysis, pathway analysis, biomarker discovery and more. The latest version, MetaboAnalyst 6.0[@pang2024], provides a unified platform for metabolomics data processing, analysis and interpretation, integrating Mummichog for pathway analysis without prior annotation.

- [PMDDA](https://yufree.github.io/pmdda/script.html) is a reproducible workflow for exhaustive MS2 data acquisition of MS1 features[@yu2022b] with data and script available online.

- [tidymass](https://www.tidymass.org/) is an object-oriented reproducible analysis framework for LC–MS data[@shen2022].

- [R for mass spectrometry](https://www.rformassspectrometry.org/) is a R software collection for the analysis and interpretation of high throughput mass spectrometry assays.

-   [MAVEN](http://genomics-pubs.princeton.edu/mzroll/index.php?show=index) from Princeton University [@melamud2010; @clasquin2012].

-   [metabolomics](https://github.com/cran/metabolomics) is a CRAN package for analysis of metabolomics data.

-   [autoGCMSDataAnal](http://software.tobaccodb.org/software/autogcmsdataanal) is a Matlab based comprehensive data analysis strategy for GC-MS-based untargeted metabolomics and [AntDAS2](http://software.tobaccodb.org/software/antdas2) provided An automatic data analysis strategy for UPLC-HRMS-based metabolomics[@yu2019; @zhang2020].

-   [enviGCMS](https://github.com/yufree/enviGCMS) from environmental non-targeted analysis and [rmwf](https://github.com/yufree/rmwf) for reproducible metabolomics workflow [@yu2020; @yu2019a].

-   Pseudotargeted metabolomics method [@zheng2020; @wang2016a].

-   [pySM](https://github.com/alexandrovteam/pySM) provides a reference implementation of our pipeline for False Discovery Rate-controlled metabolite annotation of high-resolution imaging mass spectrometry data [@palmer2017].

-   [TinyMS](https://github.com/griquelme/tidyms) is a Python-Based Pipeline for Preprocessing LC--MS Data for Untargeted Metabolomics Workflows [@riquelme2020]

-   [MetaboliteDetector](https://md.tu-bs.de/) is a QT4 based software package for the analysis of GC/MS based metabolomics data [@hiller2009].

-   [W4M](http://workflow4metabolomics.org/) and [metaX](http://metax.genomics.cn/) could analysis data online [@giacomoni2015; @wen2017; @jalili2020].

-   [FTMSVisualization](https://github.com/wkew/FTMSVisualization) is a suite of tools for visualizing complex mixture FT-MS data [@kew2017]

-   [magma](http://www.emetabolomics.org/magma) could predict and match MS/MS files.

-   [metabCombiner](https://github.com/hhabra/metabCombiner) Paired Untargeted LC-HRMS Metabolomics Feature Matching and Concatenation of Disparately Acquired Data Sets[@habra2021]

-   [SLAW](https://github.com/zamboni-lab/SLAW) is a scalable and self-Optimizing processing workflow for Untargeted LC-MS with a docker image [@delabriere2021].

-   [patRoon](https://github.com/rickhelmus/patRoon): open source software platform for environmental mass spectrometry based non-target screening [@helmus2021].

- 'shape-orientated' algorithm: A new 'shape-orientated' continuous wavelet transform (CWT)-based algorithm employing an adapted Marr wavelet (AMW) with a shape matching index (SMI), defined as peak height normalized wavelet coefficient for feature filtering, was developed for chromatographic peak detection and quantification. [@bai2022]

- automRm An R Package for Fully Automatic LC-QQQ-MS Data Preprocessing Powered by Machine Learning. [@eilertz2022] 

- [IDSL.UFA](https://ufa.idsl.me/)Intrinsic Peak Analysis (IPA) for HRMS Data. [@baygi2022]

- [DEIMoS](https://github.com/pnnl/deimos): An Open-Source Tool for Processing High-Dimensional Mass Spectrometry Data [@colby2022]

- Omics Untargeted Key Script is a tools to make untargeted LC-MS metabolomic profiling with the latest computational features readily accessible in a ready-to-use unified manner to a research community[@plyushchenko2022].

- [MetEx](http://www.metaboex.cn/MetEx) is a targeted extraction strategy for improving the coverage and accuracy of metabolite annotation[@zheng2022a].

- Asari:Trackable and scalable LC-MS metabolomics data processing software in Python[@li2023a]

- NOMspectra: An Open-Source Python Package for Processing High Resolution Mass Spectrometry Data on Natural Organic Matter[@volikov2023]

- MARS:A Multipurpose Software for Untargeted LC−MS-Based Metabolomics and Exposomics with GUI in C++ [@goracci2024]

- MeRgeION: a Multifunctional R Pipeline for Small Molecule LC-MS/MS Data Processing, Searching, and Organizing [@liu2023a]

### Workflow Comparison

Here are some comparisons for different workflow and you could make selection based on their works[@myers2017; @weber2017; @li2018a;@liao2023].

[xcmsrocker](https://github.com/yufree/xcmsrocker) is a docker image for metabolomics to compare R based software with template[@yu2022b].

## Project Setup

I suggest building your data analysis projects in RStudio (Click File - New project - New dictionary - Empty project). Then assign a name for your project. I also recommend the following tips if you are familiar with it.

-   Use [git](https://git-scm.com/)/[github](https://github.com/) to make version control of your code and sync your project online.

-   Don't use your name for your project because other peoples might cooperate with you and someone might check your data when you publish your papers. Each project should be a work for one paper or one chapter in your thesis.

-   Use **workflow** document(txt or doc) in your project to record all of the steps and code you performed for this project. Treat this document as digital version of your experiment notebook

-   Use **data** folder in your project folder for the raw data and the results you get in data analysis

-   Use **figure** folder in your project folder for the figure

-   Use **munuscript** folder in your project folder for the manuscript (you could write paper in rstudio with the help of template in [Rmarkdown](https://github.com/rstudio/rticles))

-   Just double click $$yourprojectname$$.Rproj to start your project

## Data Standards and Metadata

Reproducible metabolomics research requires not only sharing raw data but also providing well-structured metadata that describes how the data was generated and processed. The FAIR principles (Findable, Accessible, Interoperable, Reusable) provide a general framework for scientific data management[@wilkinson2016], and their adoption in metabolomics is critical for cross-study comparisons and meta-analyses.

### Metadata Standards

The Investigation-Study-Assay (ISA) framework is the most widely adopted metadata standard in metabolomics[@sansone2012]. ISA-Tab provides a structured format to describe the experimental design (Investigation), the samples and their biological context (Study), and the analytical measurements (Assay). For metabolomics specifically, mwTab is the format used by the Metabolomics Workbench[@sud2016].

When submitting data to public repositories, the minimum reporting standards proposed by the Metabolomics Standards Initiative (MSI) should be followed[@salek2013]. These standards cover the biological context, chemical analysis, data processing and statistical analysis metadata. In practice, compliance with these minimum reporting standards remains a challenge in the community[@spicer2017a].

### FAIR in Metabolomics

FAIR principles[@wilkinson2016] have been increasingly adopted in metabolomics workflows. Several tools and resources have been developed to make metabolomics data more FAIR[@rocca-serra2016]:

-   Use persistent identifiers (e.g., InChI, SMILES) for compounds and DOIs for datasets

-   Deposit raw data in open formats (mzML) to public repositories (MetaboLights[@haug2020], Metabolomics Workbench[@sud2016])

-   Document the complete analytical and computational workflow with version-controlled parameters

-   Use controlled vocabularies and ontologies (e.g., Chemical Entities of Biological Interest, ChEBI) for annotation

The metaRbolomics initiative provides an overview of R-based tools that support FAIR-compliant metabolomics workflows[@stanstrup2019]. For quality assurance and quality control standards in practice, check the mQACC consortium guidelines[@obrien2024].

### Practical Recommendations

For new metabolomics practitioners, the following checklist could help improve data quality and reproducibility[@rampler2021; @broadhurst2018a]:

-   Record all instrument parameters, column information and mobile phase composition in a machine-readable format

-   Include pooled QC and blank samples in every analytical batch and document their preparation

-   Convert vendor-specific raw files to open formats (mzML via ProteoWizard) immediately after acquisition

-   Use standardized file naming conventions that encode sample metadata (group, batch, injection order)

-   Track all data processing parameters (software version, peak picking thresholds, alignment settings) in a reproducible script or workflow file

## Data sharing

See this paper[@haug2017]:

-   [MetaboLights](http://www.ebi.ac.uk/metabolights/) EU based

-   [The Metabolomics Workbench](http://www.metabolomicsworkbench.org/) US based

-   [MetaboBank](https://mb2.ddbj.nig.ac.jp/) Japan based

-   [MetabolomeXchange](http://www.metabolomexchange.org/site/) search engine

-   [MetabolomeExpress](https://www.metabolome-express.org/) a public place to process, interpret and share GC/MS metabolomics datasets[@carroll2010].

## Contest

-   [CASMI](http://www.casmi-contest.org/) predict small molecular contest[@blazenovic2017]
