# Annotation

When you get the peaks table or features table, annotation of the peaks would help you. Check this review[@domingo-almenara2018] for a detailed notes on annotation. They proposed five levels regarding currently computational annotation strategies.

- Level 1: Peak Grouping: MS Psedospectra extraction based on peak shape similarity and peak abundance correlation

- Level 2: Peak Annotation: Adducts, Neutral losses, isotopes, and other mass relationships based on mass distances

- Level 3: Biochemical knowledge based on putative identification, potential biochemical reaction and related statistical analysis 

- Level 4: Use and intergration of tandem MS data based on data dependant/independent acquistion mode or **in silico** predction

- Level 5: Retention time prediction based on library-available retention index or quantitative structure-retnetion relationships (QSRR) models.

Most of the softwares are at level 1 or 2. If we only have compounds structure, we could guess ions under different ionization method. If we have mass spectrum, we could match the mass spectral by a similarity analysis to the database. In metabolomics, we only have mass spectrum or mass-to-charge ratios. Single mass-to-charge ratio is not enough for identification. That's the one bottleneck for annotation. So prediction is always performed on MS/MS data.

## Issues in annotation

The major issue in annotation is the redundancy peaks from same metabolite. Unlike genomcis, peaks or featuers from peak selection are not independant with each other. Adducts, in-source fragments and isotopes would lead to missannotation. A commen solution is that use known adducts, neutral losses, molecular multimers or multipley charged ions to compare mass distances.

Another issue is about the MS/MS database. Only 10% of known metabolites in databases have experimental spectral data. Thus **in silico** prediction are required. Some works try to fill the gap between experimental data, theoretical values(from chemical database like chemspider) and prediction together. Here is a nice review about MS/MS prediction[@hufsky2014].

## Annotation v.s. identification

According to the defination from the Chemical Analysis Working Group of the Metabolomics Standards Intitvative[@sumner2007;@viant2017]. Four levels of confidence could be assigned to identification:

- Level 1 'identified metabolites'
- Level 2 'Putatively annotated compounds'
- Level 3 'Putatively characterised compound classes'
- Level 4 'Unknown'

In practice, data analysis based annotation could reach level 2. For level 1, we need at extra methods such as MS/MS, retention time, accurate mass, 2D NMR spectra, and so on to confirm the compounds. However, standards are always required for solid proof.

## MS Database for annotation

### MS/MS

- [MoNA](http://mona.fiehnlab.ucdavis.edu/) Platform to collect all other open source database

- [MassBank](http://www.massbank.jp/?lang=en)

- [GNPS](https://gnps.ucsd.edu/ProteoSAFe/static/gnps-splash.jsp)  use  inner correlationship in the data and make network analysis at peaks' level instand of annotated compounds to annotate the data[@wang2016b].

- [ReSpect](http://spectra.psc.riken.jp/): phytochemicals

- [Metlin](https://metlin.scripps.edu/) is another useful online application for annotation[@guijas2018].

- [LipidBlast](http://fiehnlab.ucdavis.edu/projects/LipidBlast): in silico prediction

- [MZcloud](https://www.mzcloud.org/)

- [NIST](https://www.nist.gov/srd/nist-standard-reference-database-1a-v17): Not free

### MS

- [Fiehn Lab](http://fiehnlab.ucdavis.edu/projects/binbase-setup)

- [NIST](https://www.nist.gov/srd/nist-standard-reference-database-1a-v17): No free

- [Spectral Database for Organic Compounds, SDBS](https://sdbs.db.aist.go.jp/sdbs/cgi-bin/cre_index.cgi?lang=eng)

- [MINE](http://minedatabase.mcs.anl.gov/#/faq) is an open access database of computationally predicted enzyme promiscuity products for untargeted metabolomics

## Compounds Database

- [PubChem](https://pubchem.ncbi.nlm.nih.gov/) is an open chemistry database at the National Institutes of Health (NIH).

- [Chemspider](http://www.chemspider.com/) is a free chemical structure database providing fast text and structure search access to over 67 million structures from hundreds of data sources.

- [ChEBI](https://www.ebi.ac.uk/chebi/) is a freely available dictionary of molecular entities focused on ‘small’ chemical compounds. 

- [RefMet](http://www.metabolomicsworkbench.org/databases/refmet/index.php) A Reference list of Metabolite names.

## Software

### Adducts list

You could find adducts list [here](https://github.com/stanstrup/commonMZ) from commonMZ project.

### Molgen

[molgen](http://molgen.de) generating all structures (connectivity isomers, constitutions) that correspond to a given molecular formula, with optional further restrictions, e.g. presence or absence of particular substructures.

### Isotope

[Isotope](https://www.envipat.eawag.ch/index.php) pattern prediction

### mfFinder

[mfFinder](http://www.chemcalc.org/mf_finder/mfFinder_em_new) predict formula based on accurate mass

### CAMERA

Common [annotation](https://bioconductor.org/packages/release/bioc/html/CAMERA.html) for xcms workflow[@kuhl2012].

### RAMClustR

The software could be found [here](https://github.com/cbroeckl/RAMClustR)[@broeckling2014]. The package included a vignette as usages. Use the following code to read:

```{r RAMClustR,eval=FALSE}
vignette('RAMClustR',package = 'RAMClustR')
```

### nontarget

[nontarget](https://github.com/blosloos/nontarget) Isotope & adduct peak grouping, homologue series detection 

### xMSannotator

The software could be found [here](https://github.com/yufree/xMSannotator)[@uppal2017].

### MINE

[MINE](http://minedatabase.mcs.anl.gov/) is an open access database of computationally predicted enzyme promiscuity products for untargeted metabolomics. The annotation would be accurate for general compounds database.

### InterpretMSSpectrum

This [package](https://github.com/cran/InterpretMSSpectrum) is for annotate and interpret deconvoluted mass spectra (mass*intensity pairs) from high resolution mass spectrometry devices. You could use this package to find molecular ions for GC-MS.

### For Ident

[For-ident](https://water.for-ident.org/#!search) could give a score for identification with the help of logD(relative retention time) and/or MS/MS.

### mzmatch

Use the following code to install this package:

```{r mzmatch,eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(c("xcms", "multtest", "mzR"))
install.packages(c("rJava", "XML", "snow", "caTools",
   "bitops", "ptw", "gplots", "tcltk2"))
source ("http://puma.ibls.gla.ac.uk/mzmatch.R/install_mzmatch.R")
```

### mz.unity

You could find source code [here](https://github.com/nathaniel-mahieu/mz.unity)[@mahieu2016a] and it's for detecting and exploring complex relationships in accurate-mass mass spectrometry data.

### MAIT

You could find source code [here](https://github.com/jpgroup/MAIT)[@fernandez-albert2014a].

### ProbMetab

Provides probability ranking to candidate compounds assigned to masses, with the prior assumption of connected sample and additional previous and spectral information modeled by the user. You could find source code [here](https://github.com/rsilvabioinfo/ProbMetab)[@silva2014].

### RAMSI

You could find paper here[@baran2013].

### Sirius

[Sirius](https://bio.informatik.uni-jena.de/software/sirius/) is a new java-based software framework[@duhrkop2015] for discovering a landscape of de-novo identification of metabolites using single and tandem mass spectrometry. It could be used with [CSI:FingerID](https://www.csi-fingerid.uni-jena.de/).

### MI-Pack

You could find python software [here](http://www.biosciences-labs.bham.ac.uk/viant/mipack/)[@weber2010]

### Plantmat

[excel library](https://sourceforge.net/projects/plantmat/) based pridiction for plant metabolites[@qiu2016].

### MetFamily

[Shiny app](https://msbi.ipb-halle.de/MetFamily/) for MS and MS/MS data annotation[@treutler2016].

### Lipidmatch

[in silico](http://secim.ufl.edu/secim-tools/lipidmatch/): in silico lipid mass spectrum search[@koelmel2017].

### MolFind

JAVA based [MolFind](http://metabolomics.pharm.uconn.edu/?q=Software.html) could make annotation for unknown chemical structure by prediction based on RI, ECOM50, drift time and CID spectra[@menikarachchi2012].

### MetFusion

Java based [integration](https://github.com/mgerlich/MetFusion) of compound identiﬁcation strategies. You could access the application [here](https://msbi.ipb-halle.de/MetFusion/)[@gerlich2013].

### iMet

This online [application](http://imet.seeslab.net/) is a network-based computation method for annotation[@aguilar-mogas2017].

### Metscape

[Metscape](http://metscape.med.umich.edu/) based on Debiased Sparse Partial Correlation (DSPC) algorithm[@basu2017] to make annotation.

### MetFrag

[MetFrag](http://c-ruttkies.github.io/MetFrag/) could be used to make **in silico** prediction/match of MS/MS data[@ruttkies2016].

### LipidFrag

[LipidFrag](https://msbi.ipb-halle.de/msbi/lipidfrag)  could be used to make **in silico** prediction/match of lipid related MS/MS data[@witting2017].

### MycompoundID

[MycompoundID](http://www.mycompoundid.org/mycompoundid_IsoMS/single.jsp) could be used to search known and unknown metabolites[@li2013a] online.

### CFM-ID

[CFM-ID](https://sourceforge.net/projects/cfm-id/) use Metlin's data to make prediction[@allen2014].

### magma

[magma](http://www.emetabolomics.org/magma) could predict and match MS/MS files.

